# Exploring the Tycho Data
I'd like to start off this series of posts by taking a look at the data we're going to be working
with. As a reminder, I'm sharing the code for this series at https://github.com/dkincaid/web-app-explore.

As I mentioned last post the data that we're going to use comes from the
[Tycho project][tycho] at University of Pittsburgh. I'm going to use the
[Level 2 data][level2]. I chose this data for a couple
reasons. The first is that it is a little larger dataset (47 diseases over 1888-2013) and offers an
API to access the data.

For more information on this data see:

> Willem G. van Panhuis, John Grefenstette, Su Yon Jung, Nian Shong Chok, Anne Cross, Heather Eng,
> Bruce Y Lee, Vladimir Zadorozhny, Shawn Brown, Derek Cummings, Donald S. Burke. Contagious
> Diseases in the United States from 1888 to the present. NEJM 2013; 369(22): 2152-2158.

## Overview
I have already downloaded and processed all the data for all the states into a file named
`states_cases.Rda`. Let's load that up and have a look.

```{r loading_data}
options(width=120, warn=-1)
load("states_cases.Rda")
str(states_cases)
summary(states_cases)
```

So we have counts of the number of cases reported in each state for each week of the year. What is
the range of years?

```{r year_range}
c(min(states_cases$year), max(states_cases$year))
```
Let's see how the total cases for each disease in the data. Here I'm going to use the new package
from Hadley Wickam, [`dplyr`][dplyr] (which I am loving!).

```{r total_disease_counts}
library(dplyr, quietly=TRUE, warn.conflicts=FALSE)
states_cases %.%
    group_by(disease) %.%
    summarise(total=sum(number)) %.%
    arrange(desc(total))
```

and what about the top 10 years?

```{r total_year_counts}
states_cases %.%
    group_by(year) %.%
    summarise(total=sum(number)) %.%
    arrange(desc(total)) %.%
    head(10)
```

How about looking at the worst 10 years for each disease (I'll just show the first 3 diseases here
to save some space)

```{r topyears}
topyear <- function(d) {
    subset(states_cases, disease==d) %.%
        group_by(year) %.%
        summarise(total=sum(number)) %.%
        arrange(desc(total)) %.%
        head(10)
}

topyears.by.disease <- lapply(levels(states_cases$disease), topyear)
names(topyears.by.disease) <- levels(states_cases$disease)
head(topyears.by.disease,3)
```

So we can see that for each disease we have very different amounts of data. Let's see what the min
and max year is for each disease

```{r minmaxyear}
states_cases %.%
    group_by(disease) %.%
    summarise(min=min(year), max=max(year), range=as.numeric(max)-as.numeric(min)) %.%
    arrange(disease)
```

## Graphical visualizations
Finally, let's look at a couple of plots. Here is a look at diphtheria and whooping cough in the
states of New York, California, Texas and Illinois.

```{r, fig.height=5}
library(ggplot2, quietly=TRUE)
di <- c("DIPHTHERIA","WHOOPING COUGH [PERTUSSIS]")

dip_whoop <- subset(states_cases, disease %in% di & state %in% c("NY","CA","TX","IL"))

ggplot(dip_whoop, aes(x=as.Date(paste("1", week, year, sep="-"), format = "%w-%W-%Y"), y=number)) +
    geom_bar(stat="identity", width=7) +
    scale_y_continuous("Cases", limits=c(0,1500)) +
    scale_x_date("Week") +
    facet_grid(disease ~ state)
```
Now I'll create a matrix for Whooping Cough with states as rows and years as columns and create a
heatmap.

```{r, fig.width=10}
library(reshape2, quietly=TRUE, warn.conflicts=FALSE)
library(RColorBrewer, quietly=TRUE, warn.conflicts=FALSE)
influenza.year.by.state <- acast(subset(states_cases, disease=="INFLUENZA"), state ~ year, value.var='number', fun.aggregate=sum)
influenza.year.by.state[1:10, 1:10]
m <- melt(influenza.year.by.state)

ggplot(m, aes(x=Var2, y=Var1, fill=value)) +
    geom_tile() +
    scale_colour_brewer() +
    labs(x="Year", y="State", color="Cases") +
    theme_bw()
```

## Next time
I think that is enough for now. That should give you a decent sense for the data we'll be working
with. Next time I'll start creating the [Shiny][shiny] application that will let someone do some of
this exploration on a web based UI.

[tycho]: http://www.tycho.pitt.edu/
[level2]: http://www.tycho.pitt.edu/data/level2.php
[dplyr]: http://blog.rstudio.org/2014/01/17/introducing-dplyr
[shiny]: http://www.rstudio.com/shiny/

